---
useth---
title: "a3_mm"
format: html
editor: visual
---

Task 1 - Palmetto binary logistic regression (individual) In a separate .qmd, using the Florida palmetto data (palmetto.csv), use binary logistic regression to test feasibility of using variables plant height (height), canopy length (length), canopy width (width), and number of green leaves (green_lvs) to classify whether a palmetto is species Serenoa repens or Sabal etonia.

More information and metadata: https://portal.edirepository.org/nis/metadataviewer?packageid=edi.317.1

## Initializing the libraries

```{r}
library(tidyverse) 
library(here)
library(tidymodels)
library(rsample)

```

## Reading in the dataset

```{r}

palmetto_raw_df <- read_csv(here('data/palmetto.csv'))

palmetto_df <- palmetto_raw_df %>% 
  select(species, height, length, width, green_lvs) %>% 
  ## selecting the required variables for analysis
  mutate(species = factor(species))

levels(palmetto_df$species) ## Species 1: Serenoa repens is level 0, Species 2: Sabal etonia is Level 1

```

## Exploratory Analysis

```{r}

palmetto_df %>%
  ggplot(aes(x = height, y = length, color = species)) +
  geom_boxplot() +
  labs(x = "Plant Height", y = "Canopy Length", color = "Species") +
  theme_minimal()

# Visualization 2: Canopy Width vs Number of Green Leaves
palmetto_df %>%
  ggplot(aes(x = width, y = green_lvs, color = species)) +
  geom_point() +
  labs(x = "Canopy Width", y = "Number of Green Leaves", color = "Species") +
  theme_minimal()


```

## Testing first with basic logistic regression models

```{r}
f1 <- species ~ green_lvs + height + width + length
f2 <- species ~ green_lvs + height + width

blr1 <- glm(formula = f1, data = palmetto_df, family = binomial)
summary(blr1)

blr2 <- glm(formula = f2, data = palmetto_df, family = binomial)
summary(blr2)


```

## Testing for proportions

```{r}
palmetto_df %>%
  group_by(species) %>%
  summarize(n = n()) %>%
  ungroup() %>%
  mutate(prop = n / sum(n))

## The proportion of both the species in the dataset are almost equal at 50% each

```

## Splitting the species data for testing and training

```{r}
set.seed(35)

species_split <- initial_split(palmetto_df, prop = 0.80, strata = NULL)
species_train_df <- training(species_split)
species_test_df <- testing(species_split)
```

## Setting up Binary Logistic Regression Models: TidyModels

```{r}
blr_mdl <- logistic_reg() %>%
  set_engine('glm') ## GLM engine setup for the regression

## Testing model 1 for fit
model1_fit <- blr_mdl %>%
  fit(formula = f1, data = species_train_df)

## Testing the second model for fit
model2_fit <- blr_mdl %>%
  fit(formula = f2, data = species_train_df)

model1_fit
model2_fit
```

```{r}
species_test1_predict <- species_test_df %>%
  mutate(predict(model1_fit, new_data = species_test_df, type = 'prob'))

species_test2_predict <- species_test_df %>%
  mutate(predict(model2_fit, new_data = species_test_df, type = 'prob'))



```

Make sure you understand which species is the first ‘0’ factor level, and which is ‘1’ - you may want to convert to a factor first, then use the levels() function to check. Use repeated cross validation (ten-fold cross validation, repeated at least ten times - you can use functions from the tidymodels package to automate this, or manually perform the analysis using for-loops or purrr functions).

```{r}
set.seed(1567)
species_train_folds <- vfold_cv(species_train_df, v = 10, repeats = 10)
species_train_folds
```

```{r}

species_model1_wf <- workflow() %>%   
  add_model(blr_mdl) %>%
  add_formula(species ~ green_lvs + height + width + length)
  

species_model2_wf <- workflow() %>%   
  add_model(blr_mdl) %>%
  add_formula(species ~ green_lvs + height + width)

```

## Model 1 performs better

Based on the results of the cross validation, describe which model performs better at classification; you may wish to compare AICC and BIC values and/or area under ROC curve as well to support your decision.

```{r}
model1_fit_folds <- species_model1_wf %>%
  fit_resamples(species_train_folds)

model2_fit_folds <- species_model2_wf %>%
  fit_resamples(species_train_folds)

model1_fit_folds
model2_fit_folds

### Average the predictive performance of the ten models:
collect_metrics(model1_fit_folds)
collect_metrics(model2_fit_folds)
```

```{r}
## Pending: Train your selected model using the entire dataset, and create a finalized table (e.g., knitr::kable() and kableExtra functions) containing the binary logistic regression model results (at least coefficients, standard errors for the coefficients, and information for significance - consider using broom::tidy() to get you most of the way). 

```

```{r}
## since model 1 performs better let's classify the predictions

finalmodel_predictions <- species_test1_predict %>% 
  mutate(predicted_species = ifelse(.pred_1 >= .50, levels(palmetto_df$species)[1], levels(palmetto_df$species)[2]))

table(finalmodel_predictions %>% 
        select(species, predicted_species))



```

```{r}
library(kableExtra)

conf_matrix <- table(finalmodel_predictions$species, finalmodel_predictions$predicted_species)

# Calculate percentage of correctly classified observations
correct_percentage <- diag(conf_matrix) / rowSums(conf_matrix) * 100

# Add percentage column to the confusion matrix table
conf_matrix_with_percentage <- cbind(conf_matrix, Correct_Percentage = correct_percentage)

# Convert confusion matrix to data frame
conf_df <- as.data.frame(conf_matrix_with_percentage)

# Add row names and column names
rownames(conf_df) <- rownames(conf_matrix_with_percentage)
colnames(conf_df) <- c("Serenoa repens", "Sabal etonia", "% Correctly Classified")

# Print confusion matrix table with percentage using kableExtra
kable(conf_df, "html") %>%
  kable_styling()

```


```{r}
library(kableExtra)
library(broom)

# Get predicted values from the model
finalmodel_predictions <- augment(final_model, newdata = species_test_df)

# Calculate correct percentages
correct_percentage <- mean(finalmodel_predictions$species == finalmodel_predictions$.fitted) * 100

# Create a table with percentages using kableExtra
table_with_percentage <- finalmodel_predictions %>%
  select(species, .fitted) %>%
  mutate(Correctly_Classified = ifelse(.fitted >= 0.5, "Sabal etonia", "Serenoa repens")) %>%
  group_by(species, Correctly_Classified) %>%
  summarise(Count = n()) %>%
  mutate(Percentage = paste(round(Count / sum(Count) * 100, 2), "%")) %>%
  pivot_wider(names_from = Correctly_Classified, values_from = Count) %>%
  mutate(`% Correctly Classified` = paste0(round(correct_percentage, 2), "%")) %>%
  select(-species)

# Print confusion matrix table with percentage using kableExtra
table_with_percentage %>%
  kable("html") %>%
  kable_styling()

# Plot confusion matrix
ggplot(finalmodel_predictions, aes(x = species, y = .fitted, fill = .fitted)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.2f%%", .fitted * 100)), vjust = 1) +
  labs(x = "Actual Species", y = "Predicted Probability", fill = "Probability") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

A section that evaluates how successfully this model would “classify” a plant as the correct species, using a 50% cutoff (e.g. if the probability is \>=50% that it is species A, then it would be classified as species A). Use broom::augment() looking for the .fitted column, or predict() with type = ‘prob’, to find the probabilities (instead of log-odds) for each plant in the original dataset, then add a column for which species your model would classify that plant as (using a 50% cutoff) based on the included predictor variables. The outcome should be a finalized table showing, for each species, how many plants in the original dataset would be correctly classified and how many were incorrectly classified by the model, as well as an additional column with “% correctly classified”. Add a table caption above the table, and a 1-2 sentence conclusion paragraph after. Include an overview section with subsections that briefly summarize the dataset (this should include a well formatted data citation), the purpose of your analysis, and a pseudocode outline of the steps of your analysis. Your writeup should look and feel professional in style, tone, and substance. Optional: Consider including a photo or image that is relevant to your analysis. Consider applying a Bootswatch theme to your document (theme: themename in the Quarto document header) All code, including attached packages, should be included using code-folding. Make sure to suppress any messages & warnings. Set embed-resources to be true so your HTML is self-contained!
